---
title: An R Markdown document converted from "aggregating_R.ipynb"
output: html_document
---

Let's bring some data of COVID from Brazil:

```{r}
linkCovid1='https://github.com/DACSS-PreProcessing/DFaggregating/raw/refs/heads/main/datafiles/HIST_PAINEL_COVIDBR_2022_Parte1_20jul2022.csv'
linkCovid2='https://github.com/DACSS-PreProcessing/DFaggregating/raw/refs/heads/main/datafiles/HIST_PAINEL_COVIDBR_2022_Parte2_20jul2022.csv'
dataCovid1=read.csv(linkCovid1,sep = ';')
dataCovid2=read.csv(linkCovid2,sep = ';')
```

Let's concatenate both data frames:

```{r}
dataCovid=do.call(rbind,list(dataCovid1, dataCovid2))
```

```{r}
dataCovid$data=strptime(dataCovid$data, "%Y-%m-%d")
dataCovid$day=format(dataCovid$data,"%d")
dataCovid$year=format(dataCovid$data,"%Y")
dataCovid$month=format(dataCovid$data,"%m")
head(dataCovid)
```

```{r}
saveRDS(dataCovid,"BrasilCovid.rds")
```

Now, check the data available:

```{r}
str(dataCovid)
```

Let's keep complete data by "ESTADO":

```{r}
dataCovid=dataCovid[dataCovid$estado!="",]
```

Let's keep some columns:

```{r}

toSelect=c('regiao', 'estado', 'municipio','data', 'semanaEpi','casosNovos', 'obitosNovos')
covid=dataCovid[,toSelect]

head(covid)
```

Let's format the dates, and get date details:

```{r}
covid$data=strptime(covid$data, "%Y-%m-%d")
covid$day=format(covid$data,"%d")
covid$year=format(covid$data,"%Y")
covid$month=format(covid$data,"%m")
head(covid)
```

Let's find out about years available:

```{r}
unique(covid$year)
```

```{r}
unique(covid$month)
```

So, we have data from January to July 2022.
Let's find out: **count of new positive cases per month**:

```{r}
sum(covid$casosNovos[covid$month=='07'])
```

```{r}
sum(covid$casosNovos[covid$month=='06'])
```

```{r}
sum(covid$casosNovos[covid$month=='05'])
```

...

```{r}
sum(covid$casosNovos[covid$month=='01'])
```

We use **aggregation** to simplify the previous steps:

```{r}
# sum of cases by month
casesSumByMonth=aggregate(data=covid,casosNovos~month,sum)
casesSumByMonth
```

**AGGREGATING** capabilities allow us to produce useful output with few code:

* **The groupings**:

In the last example, _month_ was the **grouping** variable. We can have more the one of those:

```{r}
# sum of cases by estado and week
casesSumByStateAndMonth=aggregate(data=covid,casosNovos~estado + month,sum)
casesSumByStateAndMonth
```

* **The function to apply**:

We can have more than one function:

```{r}
# sum and mean of cases by estado and week
casesSumAndMeanByStateAndWeek=aggregate(data=covid,casosNovos~estado + semanaEpi,
          function(x) c(mean = mean(x), sum = sum(x) ) )


head(casesSumAndMeanByStateAndWeek,30)
```

...or better:

```{r}
casesSumAndMeanByStateAndWeek=do.call(data.frame, aggregate(data=covid,casosNovos~estado + semanaEpi,
function(x) c(mean = mean(x), sum = sum(x) ) ))
head(casesSumAndMeanByStateAndWeek,30)
```

* **The variables transformed**:

We can apply the function to more than one variable:

```{r}
# sum of cases and deaths by estado

CasesAndDeathsByState=aggregate(data=covid,
                                cbind(casosNovos,obitosNovos)~estado,
                                sum)

head(CasesAndDeathsByState,30)
```

* Function **according** to variable

The function can vary according to variable.  In this case, using **dplyr** is needed:

```{r}
library(dplyr)
covid |>
  group_by(month) |>
  summarize(casosNovos_VAR = var(casosNovos),
            casosNovos_SD = sd(casosNovos),
            obitosNovos_Median = median(obitosNovos),
            obitosNovos_Mean = mean(obitosNovos))
```

