---
title: An R Markdown document converted from "Reshaping_R.ipynb"
output: html_document
---

Let get some data:

```{r}
fragile = read.csv("https://github.com/DACSS-PreProcessing/DFops_concatenating_Py/raw/refs/heads/main/allDFs.csv")
```

A basic look:

```{r}
str(fragile)
```

We have data about countries (index of fragility and its components). The data is several years:

```{r}
table(fragile$Year)
```

Take a look:

```{r}
head(fragile)
```

What **shape** does the table have? The presence of year in a column could make us think it is in a long shape.

This could be a LONG:

```{r}
head(fragile[,c('Country','Year','Total')])
```

But this is WIDE:

```{r}
head(fragile[fragile$Year==2023,c('Country','C1_Security_Apparatus',	'C2_Factionalized_Elites',	'C3_Group_Grievance')])
```

## From Long to Wide

```{r}
fragileWide=tidyr::pivot_wider(fragile[,c('Country','Year','Total')],names_from = Year, values_from = Total,names_sort=T)
fragileWide
```

Notice that for the column to be sorted properly you need to adding *names_sort*.

The wide format is useful in several cases. In general, it looks easy.

In **base R**, you can use it directly for plotting:

```{r}
boxplot(fragileWide[,-1])
```

BUT in others such as GGPLOT, it is troublesome to use that format:

```{r}
library(ggplot2)
base=ggplot(fragileWide)
base+geom_boxplot(aes(x=as.ordered(2006),y=`2006`)) +
     geom_boxplot(aes(x=as.ordered(2010),y=`2010`)) +
     geom_boxplot(aes(x=as.ordered(2015),y=`2015`)) + labs(y='')
```

## From Wide to Long

We have  **pivot_longer** :

```{r}
fragileLong=tidyr::pivot_longer(fragileWide,!Country, names_to = "Year", values_to = "FragilityIndex")
fragileLong
```

GGPLOT will work with this:

```{r}
base = ggplot(data=fragileLong)
base + geom_boxplot(aes(x=Year,y=FragilityIndex))
```

We can also use **base R**:

```{r}
boxplot(data=fragileLong,FragilityIndex~Year)
```

This is another example without years.

Let me keep one year, and some wide-shaped columns:

```{r}
CVars_columns=c('C1_Security_Apparatus',	'C2_Factionalized_Elites',	'C3_Group_Grievance')

#only one year
fragile_CVars_wide=fragile[fragile$Year==2020,c('Country',CVars_columns)]

fragile_CVars_wide
```

```{r}
boxplot(fragile_CVars_wide[,-1])
```

```{r}
fragile_CVars_long=tidyr::pivot_longer(fragile_CVars_wide,!Country, names_to = "CVars_name", values_to = "CVars_value")
fragile_CVars_long
```

```{r}
base=ggplot(data=fragile_CVars_long)
base+geom_boxplot(aes(x=CVars_name,y=CVars_value))
```

## LONG and WIDE madness

Let me keep both long and wide elements:

```{r}
fragile_since2020=fragile[fragile$Year>=2020,c('Country','Year',CVars_columns)]
fragile_since2020
```

See this result when sending a long column into a wide format when other variables are already wide:

```{r}
fragile_since2020_wide=tidyr::pivot_wider(fragile_since2020,names_from = Year, values_from = CVars_columns)
fragile_since2020_wide
```

As you saw, the conversion into the wide format changed the column names.

```{r}
#save your work!
write.csv(fragile_since2020_wide,'fragile_since2020_wide.csv', row.names=F )
```

Now, sending the wide columns next to the long column:

```{r}
fragile_since2020_long=tidyr::pivot_longer(fragile_since2020,!c(Country,Year),names_to = "CVars_name", values_to = "CVars_value")
fragile_since2020_long
```

